An excess of positive results: Comparing the standard Psychology literature with Registered Reports

### 2020-01-13 preprint/manuscript.Rmd
minor tweaks, updated bib file

abstract: >
  When scientific studies with positive results that support the tested hypotheses have a higher probability of being published than studies with negative results, the literature will give a distorted view of the research scientists conduct. Psychology researchers have been concerned about the degree of distortion publication bias and “questionable research practices” have caused in their own literature, and developed new research and publication practices that aim to minimise these biases. Registered Reports are a new publication format in which peer review and the decision to publish take place before the study results are known. We compared the results in the full population of published Registered Reports in Psychology ($N = 71$, as of November 2018) with a random sample of hypothesis-testing studies from the standard literature ($N = 152$) by searching 633 journals for the phrase ‘test* the hypothes*’ [replicating a method by @Fanelli2010]. Analysing the first hypothesis reported in each paper, we found $96.05\%$ positive results in standard reports, but only $43.66\%$ positive results in Registered Reports. The difference remained nearly as large when direct replications were excluded from the analysis ($95.95\%$ vs $50\%$ positive results). Although our study is not a direct test of the effects of publication bias and questionable research practices on the literature, the large gap between standard reports and Registered Reports suggests that psychologists under-report negative results to an extent that threatens cumulative science.

keywords          : "Publication bias, Registered Reports"
wordcount         : "6395"

@@

```{r include = FALSE}
library("papaja")
library("bookdown")
library("rmarkdown")
library("knitr")
library("TOSTER")
library("ggplot2")
library("stringr")
library("reshape2")
library("here")
library("codebook")
```

@@

A second limitation of the current study [and of @Fanelli2010] is that standard reports were selected using the search term "test$^\ast$ the hypothes$^\ast$". 
As our results show, this phrase was virtually absent in the Registered Report population. 
The wide variety of ways to introduce a hypothesis we observed in Registered Reports suggests that a search for "test$^\ast$ the hypothes$^\ast$" will miss most of the hypothesis-testing studies in the psychological literature. 
Results based on this search phrase may not generalise to all published studies. 
For example, it is possible that authors are more likely to describe their research explicitly as a hypothesis test when they found positive results, but prefer more vague language for unsupported hypotheses (e.g., "we examined the role of ..."). 
If this were true, using other strategies to select standard reports might yield lower estimates for the positive result rate. 
However, this does not seem to be the case: 
Studies using different selection criteria for articles and hypotheses have found very similar rates of supported hypotheses. 
For example, the positive result rates in @Sterling1959, @Sterling1995, and the original studies included in the Reproducibility Project: Psychology [@OSC2015] were $97.28\%$, $95.56\%$, and $97\%$, respectively. 
@Motyl2017 report $89.17\%$ and $92.01\%$ significant results for "critical" hypothesis tests in papers published in 2003-2004 and 2013-2014, respectively. 
Therefore, although the search term used to find standard reports might limit the generalisability of our results, this search strategy seems to yield comparable estimates as the selection strategies used in different studies.

>>>>

abstract: >
  When studies with positive results that support the tested hypotheses have a higher probability of being published than studies with negative results, the literature will give a distorted view of the evidence for scientific claims. Psychological scientists have been concerned about the degree of distortion in their literature due to publication bias and inflated Type-1 error rates. Registered Reports were developed with the goal to minimise such biases: In this new publication format, peer review and the decision to publish take place before the study results are known. We compared the results in the full population of published Registered Reports in Psychology ($N = 71$ as of November 2018) with a random sample of hypothesis-testing studies from the standard literature ($N = 152$) by searching 633 journals for the phrase ‘test* the hypothes*’ [replicating a method by @Fanelli2010]. Analysing the first hypothesis reported in each paper, we found $96.05\%$ positive results in standard reports, but only $43.66\%$ positive results in Registered Reports. The difference remained nearly as large when direct replications were excluded from the analysis ($95.95\%$ vs $50\%$ positive results). This large gap suggests that psychologists under-report negative results to an extent that threatens cumulative science. Although our study did not directly test the effectiveness of Registered Reports at reducing bias, these results show that the introduction of Registered Reports has led to a much larger proportion of negative results appearing in the published literature compared to standard reports.

keywords          : "Publication bias, Registered Reports, hypothesis testing"
wordcount         : "6408"

@@

```{r include = FALSE}
library("papaja")
library("bookdown")
library("rmarkdown")
library("knitr")
library("here")
library("TOSTER")
library("ggplot2")
library("stringr")
library("reshape2")
library("codebook") # not needed to knit the manuscript, but loaded here so it automatically gets cited at the end since it was used to create the codebook
library("rio") # not needed to knit the manuscript, but loaded here so it automatically gets cited at the end since it was used to create the codebook
```

@@

A second limitation of the current study [and of @Fanelli2010] is that standard reports were selected using the search term "test$^\ast$ the hypothes$^\ast$". 
As our results show, this phrase was virtually absent in the Registered Report population. 
The wide variety of ways to introduce a hypothesis we observed in Registered Reports suggests that a search for "test$^\ast$ the hypothes$^\ast$" will miss most of the hypothesis-testing studies in the psychological literature. 
Results based on this search phrase may thus not generalise to all published studies. 
For example, it is possible that authors are more likely to describe their research explicitly as a hypothesis test when they found positive results, but prefer more vague language for unsupported hypotheses (e.g., "we examined the role of ..."). 
If this were true, using other strategies to select standard reports might yield lower estimates for the positive result rate. 
However, this does not seem to be the case: 
Studies using different selection criteria for articles and hypotheses have found very similar rates of supported hypotheses. 
For example, the positive result rates in @Sterling1959, @Sterling1995, and the original studies included in the Reproducibility Project: Psychology [@OSC2015] were $97.28\%$, $95.56\%$, and $97\%$, respectively. 
@Motyl2017 report $89.17\%$ and $92.01\%$ significant results for "critical" hypothesis tests in papers published in 2003-2004 and 2013-2014, respectively. 
Therefore, although the search term used to find standard reports might limit the generalisability of our results, this search strategy seems to yield comparable estimates as the selection strategies used in different studies.

### 2020-01-13 preprint/manuscript.tex
minor tweaks, updated bib file

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[british,,jou,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={An excess of positive results: Comparing the standard Psychology literature with Registered Reports},
            pdfkeywords={Publication bias, Registered Reports},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

% Manuscript styling
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Correspondence concerning this article should be addressed to Anne M. Scheel, Den Dolech 1, Atlas 9.417, 5600 MB, Eindhoven, The Netherlands. E-mail: a.m.scheel@tue.nl}
\keywords{Publication bias, Registered Reports\newline\indent Word count: 6395}
\usepackage{dblfloatfix}

@@

\abstract{
When scientific studies with positive results that support the tested hypotheses have a higher probability of being published than studies with negative results, the literature will give a distorted view of the research scientists conduct. Psychology researchers have been concerned about the degree of distortion publication bias and ``questionable research practices'' have caused in their own literature, and developed new research and publication practices that aim to minimise these biases. Registered Reports are a new publication format in which peer review and the decision to publish take place before the study results are known. We compared the results in the full population of published Registered Reports in Psychology (\(N = 71\), as of November 2018) with a random sample of hypothesis-testing studies from the standard literature (\(N = 152\)) by searching 633 journals for the phrase `test* the hypothes*' (replicating a method by Fanelli, 2010). Analysing the first hypothesis reported in each paper, we found \(96.05\%\) positive results in standard reports, but only \(43.66\%\) positive results in Registered Reports. The difference remained nearly as large when direct replications were excluded from the analysis (\(95.95\%\) vs \(50\%\) positive results). Although our study is not a direct test of the effects of publication bias and questionable research practices on the literature, the large gap between standard reports and Registered Reports suggests that psychologists under-report negative results to an extent that threatens cumulative science.
}

@@

In a recent commentary on benefits and challenges of open-science practices for early-career researchers, Allen \& Mehler (2019) conducted a similar investigation:
They coded the proportion of null results in the 127 biomedical and psychology Registered Reports listed in the COS database as of September 2018.
We were not aware of their parallel efforts when we planned our study in September and October 2018.
Allen and Mehler used a self-developed method to code the percentage of unsupported hypotheses in Registered Reports (counting all hypotheses in each paper) and found \(60.5\%\) unsupported hypotheses across all included Registered Reports, \(66\%\) for replication attempts, and \(54.5\%\) for novel research.
They compared these numbers to an estimate of \(5-20\%\) null results in the standard literature.
This \(5-20\%\) estimate is based on data from Fanelli (2012), who coded only the first hypothesis of each paper (identical to Fanelli, 2010), and Cristea \& Ioannidis (2018){]}, who coded the percentage of statistically significant results in figures and tables of articles published in Nature, Science, and PNAS.

@@

The positive result rate we found in standard reports (\(96.05 \%\)) is slightly higher than the \(91.5\%\) reported by Fanelli (2010), although this difference was not statistically significant.
Our replication in a more recent sample of the psychology literature thus yielded a comparably high estimate of supported hypotheses, but we cannot rule out that the positive result rate in the population has increased since 2010 (cf. Fanelli, 2012).
Furthermore, our estimate of the positive result rate for Registered Reports (\(43.66 \%\)) is comparable to the \(39.5\%\) reported by Allen \& Mehler (2019), despite some differences in method and studied population.

@@

A second limitation of the current study (and of Fanelli, 2010) is that standard reports were selected using the search term \enquote{test\(^\ast\) the hypothes\(^\ast\)}.
As our results show, this phrase was virtually absent in the Registered Report population.
The wide variety of ways to introduce a hypothesis we observed in Registered Reports suggests that a search for \enquote{test\(^\ast\) the hypothes\(^\ast\)} will miss most of the hypothesis-testing studies in the psychological literature.
Results based on this search phrase may not generalise to all published studies.
For example, it is possible that authors are more likely to describe their research explicitly as a hypothesis test when they found positive results, but prefer more vague language for unsupported hypotheses (e.g., \enquote{we examined the role of \ldots{}}).
If this were true, using other strategies to select standard reports might yield lower estimates for the positive result rate.
However, this does not seem to be the case:
Studies using different selection criteria for articles and hypotheses have found very similar rates of supported hypotheses.
For example, the positive result rates in Sterling (1959), Sterling et al. (1995), and the original studies included in the Reproducibility Project: Psychology (Collaboration, 2015) were \(97.28\%\), \(95.56\%\), and \(97\%\), respectively.
Motyl et al. (2017) report \(89.17\%\) and \(92.01\%\) significant results for \enquote{critical} hypothesis tests in papers published in 2003-2004 and 2013-2014, respectively.
Therefore, although the search term used to find standard reports might limit the generalisability of our results, this search strategy seems to yield comparable estimates as the selection strategies used in different studies.

@@

The data and code necessary to reproduce all analyses reported here, as well as additional supplementary information, is available at \url{https://osf.io/dbhgr}. The manuscript including figures and statistical analyses, the appendix, and the codebook available in the supplement were created using RStudio (1.2.5019, RStudio Team, 2019) and R (Version 3.6.0; R Core Team, 2019) and the R-packages \emph{bookdown} (Version 0.17; Xie, 2016), \emph{codebook} (Version 0.8.1; Arslan, 2018), \emph{ggplot2} (Version 3.1.1; Wickham, 2016), \emph{here} (Version 0.1; Müller, 2017), \emph{knitr} (Version 1.26; Xie, 2015), \emph{papaja} (Version 0.1.0.9842; Aust \& Barth, 2018), \emph{reshape2} (Version 1.4.3; Wickham, 2007), \emph{rmarkdown} (Version 2.0; Xie, Allaire, \& Grolemund, 2018), \emph{stringr} (Version 1.4.0; Wickham, 2019), and \emph{TOSTER} (Version 0.3.4; Lakens, 2017).

@@

\leavevmode\hypertarget{ref-OSC2015}{}%
Collaboration, O. S. (2015). Estimating the reproducibility of psychological science. \emph{Science}, \emph{349}(6251), aac4716. \url{https://doi.org/10.1126/science.aac4716}

@@

\leavevmode\hypertarget{ref-Agnoli2017}{}%
Agnoli, F., Wicherts, J. M., Veldkamp, C. L. S., Albiero, P., \& Cubelli, R. (2017). Questionable research practices among italian research psychologists. \emph{PLOS ONE}, \emph{12}(3), e0172792. \url{https://doi.org/10.1371/journal.pone.0172792}

\leavevmode\hypertarget{ref-OSC2015}{}%
Collaboration, O. S. (2015). Estimating the reproducibility of psychological science. \emph{Science}, \emph{349}(6251), aac4716. \url{https://doi.org/10.1126/science.aac4716}

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[british,,jou,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={An excess of positive results: Comparing the standard Psychology literature with Registered Reports},
            pdfkeywords={Publication bias, Registered Reports, hypothesis testing},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
% Manuscript styling
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

Correspondence concerning this article should be addressed to Anne M. Scheel, Den Dolech 1, Atlas 9.417, 5600 MB, Eindhoven, The Netherlands. E-mail: a.m.scheel@tue.nl}
\keywords{Publication bias, Registered Reports, hypothesis testing\newline\indent Word count: 6408}
\usepackage{dblfloatfix}

@@

\abstract{
When studies with positive results that support the tested hypotheses have a higher probability of being published than studies with negative results, the literature will give a distorted view of the evidence for scientific claims. Psychological scientists have been concerned about the degree of distortion in their literature due to publication bias and inflated Type-1 error rates. Registered Reports were developed with the goal to minimise such biases: In this new publication format, peer review and the decision to publish take place before the study results are known. We compared the results in the full population of published Registered Reports in Psychology (\(N = 71\) as of November 2018) with a random sample of hypothesis-testing studies from the standard literature (\(N = 152\)) by searching 633 journals for the phrase `test* the hypothes*' (replicating a method by Fanelli, 2010). Analysing the first hypothesis reported in each paper, we found \(96.05\%\) positive results in standard reports, but only \(43.66\%\) positive results in Registered Reports. The difference remained nearly as large when direct replications were excluded from the analysis (\(95.95\%\) vs \(50\%\) positive results). This large gap suggests that psychologists under-report negative results to an extent that threatens cumulative science. Although our study did not directly test the effectiveness of Registered Reports at reducing bias, these results show that the introduction of Registered Reports has led to a much larger proportion of negative results appearing in the published literature compared to standard reports.
}

@@

In a recent commentary on benefits and challenges of open-science practices for early-career researchers, Allen and Mehler (2019) conducted a similar investigation:
They coded the proportion of null results in the 127 biomedical and psychology Registered Reports listed in the COS database as of September 2018.
We were not aware of their parallel efforts when we planned our study in September and October 2018.
Allen and Mehler used a self-developed method to code the percentage of unsupported hypotheses in Registered Reports (counting all hypotheses in each paper) and found \(60.5\%\) unsupported hypotheses across all included Registered Reports, \(66\%\) for replication attempts, and \(54.5\%\) for novel research.
They compared these numbers to an estimate of \(5-20\%\) null results in the standard literature.
This \(5-20\%\) estimate is based on data from Fanelli (2012), who coded only the first hypothesis of each paper (identical to Fanelli, 2010), and Cristea and Ioannidis (2018){]}, who coded the percentage of statistically significant results in figures and tables of articles published in Nature, Science, and PNAS.

@@

The positive result rate we found in standard reports (\(96.05 \%\)) is slightly higher than the \(91.5\%\) reported by Fanelli (2010), although this difference was not statistically significant.
Our replication in a more recent sample of the psychology literature thus yielded a comparably high estimate of supported hypotheses, but we cannot rule out that the positive result rate in the population has increased since 2010 (cf. Fanelli, 2012).
Furthermore, our estimate of the positive result rate for Registered Reports (\(43.66 \%\)) is comparable to the \(39.5\%\) reported by Allen and Mehler (2019), despite some differences in method and studied population.

@@

A second limitation of the current study (and of Fanelli, 2010) is that standard reports were selected using the search term \enquote{test\(^\ast\) the hypothes\(^\ast\)}.
As our results show, this phrase was virtually absent in the Registered Report population.
The wide variety of ways to introduce a hypothesis we observed in Registered Reports suggests that a search for \enquote{test\(^\ast\) the hypothes\(^\ast\)} will miss most of the hypothesis-testing studies in the psychological literature.
Results based on this search phrase may thus not generalise to all published studies.
For example, it is possible that authors are more likely to describe their research explicitly as a hypothesis test when they found positive results, but prefer more vague language for unsupported hypotheses (e.g., \enquote{we examined the role of \ldots{}}).
If this were true, using other strategies to select standard reports might yield lower estimates for the positive result rate.
However, this does not seem to be the case:
Studies using different selection criteria for articles and hypotheses have found very similar rates of supported hypotheses.
For example, the positive result rates in Sterling (1959), Sterling et al. (1995), and the original studies included in the Reproducibility Project: Psychology (Open Science Collaboration, 2015) were \(97.28\%\), \(95.56\%\), and \(97\%\), respectively.
Motyl et al. (2017) report \(89.17\%\) and \(92.01\%\) significant results for \enquote{critical} hypothesis tests in papers published in 2003-2004 and 2013-2014, respectively.
Therefore, although the search term used to find standard reports might limit the generalisability of our results, this search strategy seems to yield comparable estimates as the selection strategies used in different studies.

@@

The data and code necessary to reproduce all analyses reported here, as well as additional supplementary information, is available at \url{https://osf.io/dbhgr}. The manuscript including figures and statistical analyses, the appendix, and the codebook available in the supplement were created using RStudio (1.2.5019, RStudio Team, 2019) and R (Version 3.6.0; R Core Team, 2019) and the R-packages \emph{bookdown} (Version 0.17; Xie, 2016), \emph{codebook} (Version 0.8.2; Arslan, 2018), \emph{ggplot2} (Version 3.1.1; Wickham, 2016), \emph{here} (Version 0.1; Müller, 2017), \emph{knitr} (Version 1.26; Xie, 2015), \emph{papaja} (Version 0.1.0.9842; Aust \& Barth, 2018), \emph{reshape2} (Version 1.4.3; Wickham, 2007), \emph{rio} (Version 0.5.16; Chan, Chan, Leeper, \& Becker, 2018), \emph{rmarkdown} (Version 1.18; Xie, Allaire, \& Grolemund, 2018), \emph{stringr} (Version 1.4.0; Wickham, 2019), and \emph{TOSTER} (Version 0.3.4; Lakens, 2017).

@@

\leavevmode\hypertarget{ref-R-rio}{}%
Chan, C.-h., Chan, G. C., Leeper, T. J., \& Becker, J. (2018). \emph{Rio: A swiss-army knife for data file i/o}.

@@

\leavevmode\hypertarget{ref-OSC2015}{}%
Open Science Collaboration. (2015). Estimating the reproducibility of psychological science. \emph{Science}, \emph{349}(6251), aac4716. \url{https://doi.org/10.1126/science.aac4716}
